Temporal ordering and registration of images in studies of developmental dynamics

### 2014-06-02 image_analysis_paper/si_information.tex
Added yannis edits before sending to Amit

Each image is subsampled to $100 \times 100$ pixels.
%
For the first and third data sets in the paper (images taken during cellularization), we only use the red and green channels (dpERK and Dl) in the vector diffusion maps computations, as the spatial location of the nuclei does not change appreciably during this time window.
%
The nuclei in the corresponding figures are shown only for visual reference.
%
For the second data set of images during gastrulation, we use all three channels (dpERK, Twi, and nuclei) in the vector diffusion maps computations.
%
We blur the nuclear signal using a disc filter with a $5$~pixel radius (using the \texttt{imfilter} function in Matlab with a disc filter).
%
We also adjust the intensity of the nuclear signal in each image such that 1\% of the pixels are saturated (using the \texttt{imadjust} function in Matlab).

When imaging, the laser settings were calibrated so that images have good dynamic range, while not saturating the detector. 
% 
The same laser setting was maintained within each experiment.

\subsection{Mathematical Algorithms}

We will demonstrate the algorithms for registration and temporal ordering using a synthetic data set.
%
The relatively simple dynamics of this data set will allow us to easily visualize and illustrate the main points of the different algorithms.
%
We construct concentration profiles defined on a ring, and we assume that each ring is arbitrarily rotated around its center; an example is shown in \fig~\ref{subfig:1d_example}.
%
Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown at the bottom of \fig~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
%
Each concentration profile is a noisy Gaussian (shown in \fig~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
%
We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_{100}$, the group of integers modulo $100$).
%
\fig~\ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.

@@

Let $ x_1, \dots, x_m$ denote the signals that we wish to align with respect to rotations;
each signal is a function defined on the unit circle (on the plane).
%
%Practically, it is discretized in a $n$-long vector (the local intensity at $n$ equidistant points around the circle);
%rotating the function by an angle $\theta$ then corresponds to cyclically shifting the elements of $x_i$ by $\frac{\theta_i}{2 \pi} n$ (rounded to the nearest integer to obtain a valid shift).
%
First assume that each signal $x_i$ is a {\it noisy} rotated copy of the underlying signal $x_{true}$
(which we are {\it not} given), such that
\begin{equation}
x_i = f(x_{true}, \theta_i) + \xi_i
\end{equation}
where the function $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ radians, and $\xi_i$ is a (typically Gaussian) noise term.
%
Our goal is to recover $\theta_1, \dots, \theta_m$.
%
Up to noise,
\begin{equation} \label{eq:pairwise_rot}
x_i \approx f(x_j, \theta_i - \theta_j) ;
\end{equation}
note that \eqref{eq:pairwise_rot} does not require knowledge of $x_{true}$.
%
We can obtain an {\it estimate} of $\theta_i - \theta_j$ by computing the rotation that optimally aligns $x_j$ to $x_i$,
i.e., %$\theta_{ij} \approx \theta_i - \theta_j$, where
%
\begin{equation} \label{eq:opt_angle}
\theta_i - \theta_j \approx \theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2.
\end{equation}
%
Practically, the signals are discretized in a $n$-long vector (the local intensity at $n$ equidistant points around the circle);
rotating the function by an angle $\theta$ then corresponds to cyclically shifting the elements of $x_i$
by $\frac{\theta_i}{2 \pi} n$ (rounded to the nearest integer to obtain a valid shift).
%
For the one-dimensional discretized profiles shown in \fig~\ref{fig:1d_demo}, we exhaustively search over all $n=100$ possible shifts of the signals to compute the optimal angles in \eqref{eq:opt_angle}.
%
Alternatively, for continuous signals, an optimization algorithm
can be used \cite{ahuja2007template}.

@@

Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$; we write our procedure for general $d$ because we will later consider three-dimensional rotations).
%
We construct the matrix $H \in \mathbb{R}^{md \times md}$, where $H$ is an $m \times m$ matrix of $d \times d$ blocks, with the $i,j^{th}$ block of $H$, $H_{ij}$, defined as
\begin{equation} \label{eq:H_to_R}
H_{ij} = R(\theta_{ij}).
\end{equation}
%
%
Under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, $H_{ij} \approx R(\theta_i) R(\theta_j)^T$
%\begin{equation}
%H_{ij} = R(\theta_{ij}) \approx R(\theta_i - \theta_j) = R(\theta_i) R(-\theta_j) = R(\theta_i) R(\theta_j)^T,
%\end{equation}
 and
\begin{equation} \label{eq:H_low_rank}
	H \approx
	\begin{bmatrix}
	R(\theta_1) \\
	R(\theta_2) \\
	\vdots \\
	R(\theta_m)
	\end{bmatrix}
	\begin{bmatrix}
	R(\theta_1)^T R(\theta_2)^T \dots R(\theta_m)^T
	\end{bmatrix}.
\end{equation}
%
It follows directly from \eqref{eq:H_low_rank} that the top block eigenvector of $H$ contains our best estimates of $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$.
%
Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$.
%
Then,
\begin{equation} \label{eq:R_hat}
\hat{R} =
\begin{bmatrix}
\hat{R}_1 \\
\hat{R}_2 \\
\vdots \\
\hat{R}_m
\end{bmatrix} =
\begin{bmatrix}
| & | & & | \\
\phi_1 & \phi_2 & \dots & \phi_d \\
| & | & & |
\end{bmatrix},
\end{equation}
where $\hat{R}_i \in \mathbb{R}^{d \times d}$ is (nearly) the estimate for $R(\theta_i)$.
%
To obtain our estimate of $R(\theta_i$), denoted $R_{i, est}$, we project $\hat{R}_i$ onto the closest orthogonal matrix, 
\begin{equation} \label{eq:R_est}
R_{i, est} = U_i V_i^T,
\end{equation}
where $U_i$ and $V_i$ are the left and right singular vectors, respectively, of $\hat{R}_i$.
%
We adjust the sign of $\phi_1$ so that $det(R_{i, est}) = +1$, ensuring proper rotations.
%
We estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and register the signals by rotating signal $i$ by $-\theta_i$.
%
We note that, in our actual computations, the pairwise rotations $\theta_{ij}$ are computed in a discrete setting, then the overall
synchronization is performed in the continuum context to obtain $\theta_i$, and the results are rounded to give the closest
discrete shift.

Furthermore, this formulation also considers {\it higher-order} consistency information.
%
For example, given our pairwise estimates $R_{ij}$, we know that relationships of the form
\begin{equation} \label{eq:triplet_consistency}
R(\theta_{ik}) R(\theta_{kj}) \approx R(\theta_i) R(\theta_k)^T R(\theta_k) R(\theta_j)^T = R(\theta_i) R(\theta_j)^T
\end{equation}
should also hold.
%
Note that
\begin{equation}
(H^2)_{ij} = \sum_k R(\theta_{ik}) R(\theta_{kj});
\end{equation}
therefore, {\it all} information of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$.
%
Because $H$ and $H^2$ have the same eigenvectors, our problem formulation accounts for not only pairwise alignment information, but also these higher-order considerations.

@@

Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional space), we want to find a coordinate transformation $y(x)$ that preserves local information: points that are close in the original space should also be close in the coordinate $y$.
%
The first step is to construct the matrix $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if points $x_i$ and $x_j$ are ``close.''
%
We use a diffusion kernel,
\begin{equation} \label{eq:dmaps_W}
W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right),
%W_{ij} = e^{ -\frac{d^2(x_i, x_j)}{\epsilon^2}}
\end{equation}
where $d(x_i, x_j)$ is a pairwise distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic scale.
%
Points less than $\epsilon$ apart are thus considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
%
$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/10^{th}$ of the median of the pairwise distances between data points for the first two data sets, and half of the median of the pairwise distances between data points for the third data set.

To find this coordinate $y$, we want solve the following optimization problem \cite{Belkin2003}
\begin{equation} \label{eq:dmaps_opt_problem}
\argmin_{y} \sum_{ij} W_{ij} (y(x_i) - y(x_j))^2.
\end{equation}
%
We first compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{m} W_{ij}$, and the matrix $A$, where
\begin{equation} \label{eq:dmaps_A}
A = D^{-1} W.
\end{equation}
%
We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_m$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$.
%
Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors.
%
Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector; this is a trivial solution to \eqref{eq:dmaps_opt_problem}.
%
%In general, the next few eigenvectors $\phi_2, \dots, \phi_m$ give ``meaningful'' embedding coordinates for the data, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point.
%
The next eigenvector, $\phi_2$, is the (non-trivial) solution to \eqref{eq:dmaps_opt_problem}, so that $\phi_2(j)$, the $j^{th}$ entry of $\phi_2$, gives the ``new" coordinate for data point $x_j$ (i.e., $\phi_2(j) = y(x_j)$).
%
In our application, we have assumed that this single direction of variability, parameterized by $\phi_2$, is one-to-one with time.
%
Ordering the data by $\phi_2(j)$ will then, effectively, order them in time.
%
The procedure generalizes when the data lie on higher-dimensional manifolds (not just curves) in data space, where lower eigenvectors give subsequent embedding coordinates for the data.
%

@@

In vector diffusion maps, given data points $x_1, \dots, x_m$, one first constructs the matrix $S \in \mathbb{R}^{md \times md}$, with the $i,j^{th}$ block of $S$, $S_{ij}$, defined as
\begin{equation} \label{eq:vdm_S}
	S_{ij} = A_{ij} H_{ij}
\end{equation}
%
where $A_{ij} \in \mathbb{R}$ (defined in \eqref{eq:dmaps_A}) pertains to the diffusion kernel between data points, and $H_{ij} \in \mathbb{R}^{d \times d}$ (defined in \eqref{eq:H_to_R}) pertains to the pairwise alignment between data points.
%
It is important to note that distance $d(x_i, x_j)$ used in the diffusion kernel in \eqref{eq:dmaps_W} is the distance between data points {\it after} after pairwise alignment, i.e., the minimum distance between all possible shifts of the two data points (which is obtained in \eqref{eq:opt_pairwise}).
%
In the language of symmetry groups, this distance is a metric between the orbits induced by the symmetry group.

@@

In general, the embedding coordinates are given by
\begin{equation} \label{eq:vdm_coord}
\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle,
\end{equation}
where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$, 
%
If we assume that the rotations and the dynamics are uncoupled and therefore separable, then the eigenvectors of $S$ have the following structure: each block eigenvector contains estimates of the optimal rotations (up to a constant rotation) multiplied by the corresponding embedding coordinate.
%
As the first diffusion maps coordinate is constant, the first block eigenvector contains only the optimal rotations.
%
The second block eigenvector (eigenvectors $d+1$ through $2d$) contains the optimal rotations, each multiplied by their second diffusion maps coordinate. 
%
We can therefore recover this diffusion maps coordinate by taking inner products of the columns of the second block eigenvector with columns of the first block eigenvector.
%
The $j^{th}$ embedding coordinate will be given by $\psi_{k,1}$, where $jd +1 < k \le (j+1)d$,
and we select $k$ such that the coordinate $\psi_{k, 1}$ has the largest variability, i.e., the $j^{th}$ coordinate is $\psi_{k,1}$, where $k$ is the solution to
$$ \max_{jd +1 < k \le (j+1)d} \sum_i \psi_{k,1} (i)^2. $$

@@

Practically, we have square images discretized as pixels (rather than continuous functions on the plane).

@@

The translations are implemented by shifting the pixels;

@@

The synchronization formulation requires that the matrices used to represent the symmetry group be orthogonal \cite{singer2013spectral}.  
%(for \eqref{eq:H_low_rank} to be satisfied).
%
However, the typical matrix representation for $ISO(2)$, the group of two-dimensional translations and rotations, does not satisfy this property.
%
Instead, we {\it approximately} represent rotations and translations in two dimensions using {\it three-dimensional rotation matrices}, by projecting the (two-dimensional) image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular}.
%
Rotation of the image corresponds to rotation around one principal axis in three dimensions, while translation of the image corresponds (approximately) to rotations around the other two principal axes. (Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$. However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.)
%
These rotation matrices are (by definition) orthogonal, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.

>>>>

The original resolution of each image is $1024 \times 1024$ pixels.
%
The images are subsampled to $100 \times 100$ pixels for vector diffusion maps analysis.
%
A scale bar for a representative image is shown in \fig~\ref{fig:ap_dv}.
%
When imaging, the laser settings are calibrated so that images have good dynamic range, while not saturating the detector.
%
The same laser settings are maintained within each experiment.
%
The intensity of each channel is integer-valued, and ranges from $0$ (no signal) to $255$ (saturated). 
%
We blur the nuclear signal using a disc filter with a $5$~pixel radius (using the \texttt{imfilter} function in MATLAB\textsuperscript{\textregistered}).
%
We also adjust the intensity of the nuclear signal in each image such that 1\% of the pixels are saturated 
(using the \texttt{imadjust} function in MATLAB), and then scale the intensity of the nuclear signal by half.

For the first and third data sets in the paper (images taken during cellularization), we only use the red and green channels (dpERK and Dl) in the vector diffusion maps computations, as the spatial location of the nuclei (the gray channel) does not change appreciably during this time window.
%
The nuclei in the corresponding figures are thus shown only for visual reference.
%
For the second data set of images (taken during gastrulation), we use all three channels (dpERK, Twi, and nuclei) in the vector diffusion maps computations.
%
For this particular data set, we multiply the intensity of the Twi signal by 1.5, as this channel was not completely saturated in the original imaging data.

\subsection{Algorithms}

We will demonstrate the algorithms for registration and temporal ordering using a synthetic data set.
%
The relatively simple dynamics of this data set will allow us to easily visualize and illustrate the main features of the different algorithms.
%
Motivated by the geometry of our images, we construct a sequence of concentration profiles defined on a ring, and rotate each ring randomly around its center; an example is shown in \fig~\ref{subfig:1d_example}.
%
Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown at the bottom of \fig~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
%
Each concentration profile is a noisy Gaussian (shown in \fig~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as a function of ``time".
%
We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_{100}$, the group of integers modulo $100$).
%
\fig~\ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.

@@

Let $ x_1, \dots, x_m$ denote the signals that we wish to align with respect to rotations;
each signal is a function defined on the unit circle (on the plane).
%
%Practically, it is discretized in a $n$-long vector (the local intensity at $n$ equidistant points around the circle);
%rotating the function by an angle $\theta$ then corresponds to cyclically shifting the elements of $x_i$ by $\frac{\theta_i}{2 \pi} n$ (rounded to the nearest integer to obtain a valid shift).
%
First assume that each signal $x_i$ is a {\it noisy} rotated copy of the underlying signal $x_{true}$
(which we are {\it not} given), such that
\begin{equation}
x_i = f(x_{true}, \theta_i) + \xi_i
\end{equation}
where the function $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ radians, and $\xi_i$ is a (typically Gaussian) noise term.
%
Our goal is to recover $\theta_1, \dots, \theta_m$.
%
Up to noise,
\begin{equation} \label{eq:pairwise_rot}
x_i \approx f(x_j, \theta_i - \theta_j) ;
\end{equation}
note that \eqref{eq:pairwise_rot} does not require knowledge of $x_{true}$.
%
We can obtain an {\it estimate} of $\theta_i - \theta_j$ by computing the rotation that optimally aligns $x_j$ to $x_i$,
i.e., %$\theta_{ij} \approx \theta_i - \theta_j$, where
%
\begin{equation} \label{eq:opt_angle}
\theta_i - \theta_j \approx \theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2.
\end{equation}
%
Practically, the signals are discretized in a $n$-long vector (the local intensity at $n$ equidistant points around the circle);
rotating the function by an angle $\theta$ then corresponds to cyclically shifting the elements of $x_i$
by $\frac{\theta_i}{2 \pi} n$ (rounded to the nearest integer to obtain a valid shift).
%
For the one-dimensional discretized profiles shown in \fig~\ref{fig:1d_demo}, we exhaustively search over all $n=100$ possible shifts of the signals to obtain the optimal angles in \eqref{eq:opt_angle}.
%
Alternatively, for continuous signals, an optimization algorithm
can be used \cite{ahuja2007template}.

@@

Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$; we write our procedure for general $d$ because we will later consider three-dimensional rotations).
%
We construct the matrix $H \in \mathbb{R}^{md \times md}$, where $H$ is an $m \times m$ matrix of $d \times d$ blocks, with the $i,j^{th}$ block of $H$, $H_{ij}$, defined as
\begin{equation} \label{eq:H_to_R}
H_{ij} = R(\theta_{ij}).
\end{equation}
%
%
Under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, $H_{ij} \approx R(\theta_i) R(\theta_j)^T$
%\begin{equation}
%H_{ij} = R(\theta_{ij}) \approx R(\theta_i - \theta_j) = R(\theta_i) R(-\theta_j) = R(\theta_i) R(\theta_j)^T,
%\end{equation}
 and
\begin{equation} \label{eq:H_low_rank}
	H \approx
	\begin{bmatrix}
	R(\theta_1) \\
	R(\theta_2) \\
	\vdots \\
	R(\theta_m)
	\end{bmatrix}
	\begin{bmatrix}
	R(\theta_1)^T R(\theta_2)^T \dots R(\theta_m)^T
	\end{bmatrix}.
\end{equation}
%
It follows directly from \eqref{eq:H_low_rank} that the top block eigenvector of $H$ contains our best estimates of $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$.
%
Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$ ordered so that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$.
%
Then,
\begin{equation} \label{eq:R_hat}
\hat{R} =
\begin{bmatrix}
\hat{R}_1 \\
\hat{R}_2 \\
\vdots \\
\hat{R}_m
\end{bmatrix} =
\begin{bmatrix}
| & | & & | \\
\phi_1 & \phi_2 & \dots & \phi_d \\
| & | & & |
\end{bmatrix},
\end{equation}
where $\hat{R}_i \in \mathbb{R}^{d \times d}$ is (nearly) the estimate for $R(\theta_i)$.
%
To obtain our estimate of $R(\theta_i$), denoted $R_{i, est}$, we project $\hat{R}_i$ onto the closest orthogonal matrix,
\begin{equation} \label{eq:R_est}
R_{i, est} = U_i V_i^T,
\end{equation}
where $U_i$ and $V_i$ are the left and right singular vectors, respectively, of $\hat{R}_i$.
%
We adjust the sign of $\phi_1$ so that $det(R_{i, est}) = +1$, ensuring proper rotations 
(note that systematically incorporating improper rotations is also possible \cite{goemans1995improved, bandeira2013cheeger}).
%
We estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and register the signals by rotating signal $i$ by $-\theta_i$.
%
We note that, in our actual computations, the pairwise rotations $\theta_{ij}$ are computed in a discrete setting, then the overall
synchronization is performed in the continuum context to obtain $\theta_i$, and the results are rounded to give the closest
discrete shift.

Importantly, this formulation also considers {\it higher-order} consistency information.
%
For example, given our pairwise estimates $R_{ij}$, we know that relationships of the form
\begin{equation} \label{eq:triplet_consistency}
R(\theta_{ik}) R(\theta_{kj}) \approx R(\theta_i) R(\theta_k)^T R(\theta_k) R(\theta_j)^T = R(\theta_i) R(\theta_j)^T
\end{equation}
should also hold.
%
Note that
\begin{equation}
(H^2)_{ij} = \sum_k R(\theta_{ik}) R(\theta_{kj});
\end{equation}
therefore, {\it all} information of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$ (and higher order
consistency information in its higher powers).
%
Because $H$ and $H^2$ have the same eigenvectors, our problem formulation accounts for not only pairwise alignment information, but also these higher-order considerations.

@@

Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional vector space), we want to find a coordinate transformation $y(x)$ that preserves local geometry: points that are ``close" in the original space should also be ``close" in the coordinates $y$.
%
The first step is to construct the matrix $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if points $x_i$ and $x_j$ are ``close.''
%
We use a diffusion kernel,
\begin{equation} \label{eq:dmaps_W}
W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right),
%W_{ij} = e^{ -\frac{d^2(x_i, x_j)}{\epsilon^2}}
\end{equation}
where $d(x_i, x_j)$ is a pairwise distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic scale.
%
Points less than $\epsilon$ apart are thus considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
%
$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/10^{th}$ of the median of the pairwise distances between data points for the first two data sets (and half of the median of the pairwise distances between data points for the third data set).

To find the coordinate $y$, we want solve the following optimization problem \cite{Belkin2003}
\begin{equation} \label{eq:dmaps_opt_problem}
\argmin_{y} \sum_{ij} W_{ij} (y(x_i) - y(x_j))^2.
\end{equation}
%
We first compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{m} W_{ij}$, and the matrix $A$, where
\begin{equation} \label{eq:dmaps_A}
A = D^{-1} W.
\end{equation}
%
We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_m$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$.
%
Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors.
%
Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector; this is a trivial solution to \eqref{eq:dmaps_opt_problem}.
%
%In general, the next few eigenvectors $\phi_2, \dots, \phi_m$ give ``meaningful'' embedding coordinates for the data, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point.
%
The next eigenvector, $\phi_2$, is the (non-trivial) solution to \eqref{eq:dmaps_opt_problem}, so that $\phi_2(j)$, the $j^{th}$ entry of $\phi_2$, gives the ``new" coordinate for data point $x_j$ (i.e., $\phi_2(j) = y(x_j)$).
%
In our application, we have assumed that this {\em single} direction of variability, parameterized by $\phi_2$, is one-to-one with time.
%
Ordering the data by $\phi_2(j)$ will then, effectively, order them in time.
%
The procedure generalizes when the data lie on higher-dimensional manifolds (not just curves) in data space, where leading eigenvectors can give subsequent embedding coordinates for the data.
%

@@

In vector diffusion maps, given data points $x_1, \dots, x_m$, one first constructs the matrix $S \in \mathbb{R}^{md \times md}$, with the $i,j^{th}$ block of $S$, $S_{ij}$, defined as
\begin{equation} \label{eq:vdm_S}
	S_{ij} = A_{ij} H_{ij}
\end{equation}
%
where $A_{ij} \in \mathbb{R}$ (defined in \eqref{eq:dmaps_A}) pertains to the diffusion kernel between data points, and $H_{ij} \in \mathbb{R}^{d \times d}$ (defined in \eqref{eq:H_to_R}) pertains to the pairwise alignment between data points.
%
It is important to note that distance $d(x_i, x_j)$ used in the diffusion kernel in \eqref{eq:dmaps_W} is the distance between data points {\it after} after pairwise alignment, i.e., the minimum distance between all possible shifts of the two data points (which is obtained in \eqref{eq:opt_pairwise}).
%
In the language of symmetry groups, this distance is a metric between the orbits induced by the relevant symmetry group.

@@

In general, the embedding coordinates are given by
\begin{equation} \label{eq:vdm_coord}
\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle,
\end{equation}
where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$,
%
If we assume that the rotations and the dynamics are uncoupled and therefore separable, then the eigenvectors of $S$ have the following structure: each block eigenvector contains estimates of the optimal rotations (up to a constant rotation) multiplied by the corresponding embedding coordinate (a scalar)
%
As the first diffusion maps coordinate is constant over the data, the first block eigenvector contains only the optimal rotations.
%
The second block eigenvector (eigenvectors $d+1$ through $2d$) contains the optimal rotations, each multiplied by their second diffusion maps coordinate.
%
We can therefore recover this diffusion maps coordinate by taking inner products of the columns of the second block eigenvector with columns of the first block eigenvector.
%%
%%YGK just as a favor to me, tell me how consistent the three values are -- 
%%
The $j^{th}$ embedding coordinate will be given by $\psi_{k,1}$, where $jd +1 < k \le (j+1)d$,
and we select $k$ such that the coordinate $\psi_{k, 1}$ has the largest variability, i.e., the $j^{th}$ coordinate is $\psi_{k,1}$, where $k$ is the solution to
$$ \max_{jd +1 < k \le (j+1)d} \sum_i \psi_{k,1} (i)^2. $$

@@

Practically, we have square images discretized as pixels (rather than continuous functions on the plane); note that the center of the
image square is {\em not} necessarily also the center of the embryo.

@@

The vertical and horizontal translations are implemented by shifting the pixels;

@@

The synchronization formulation requires the matrices used to represent the symmetry group to be orthogonal \cite{singer2013spectral}.
%(for \eqref{eq:H_low_rank} to be satisfied).
%
However, the typical matrix representation for $ISO(2)$, the group of two-dimensional translations and rotations, does not satisfy this property.
%
Instead, we {\it approximately} represent rotations and translations in two dimensions using {\it three-dimensional rotation matrices}, by projecting the (two-dimensional) image onto a portion of the surface of a (three-dimensional) sphere \cite{singer2011angular}.
%
Rotation of the image corresponds to rotation around one principal axis in three dimensions, while translation of the image corresponds (approximately) to rotations around the other two principal axes (see \fig~\ref{fig:SO3_picture}). (Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$. However, the images we will be considering are already ``mostly centered", and so we are only interested in {\em small} translations that are well within the $[0, 2\pi)$ range.)
%
These rotation matrices are (by definition) orthogonal, and successive application of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.

To convert the optimal two-dimensional translations and rotation to corresponding (continuous) three-dimensional rotations,
we first compute the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$,
\begin{equation} \label{eq:angle_relations}
%\begin{aligned}
%	\alpha_{ij} &= \theta_{ij} \\
%	\beta_{ij} &= \frac{dx_{ij}}{n} \times \eta_{proj} \\
%	\gamma_{ij} &= \frac{dy_{ij}}{n} \times \eta_{proj} \\
%\end{aligned}
	\alpha_{ij} = \theta_{ij}, \; \; \;
	\beta_{ij} = \frac{dx_{ij}}{n} \times \eta_{proj}, \; \; \;
	\gamma_{ij} = \frac{dy_{ij}}{n} \times \eta_{proj}
\end{equation}
where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project the image.
%
We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere;
this portion of the sphere is small enough so that curvature effects are not prominent in the image.
%
We write rotations around the three principal axes, $R^x(\alpha)$, $R^y(\beta)$, and $R^z(\gamma)$, in terms of the three Euler angles
%\small
\begin{equation}
\begin{aligned}
	R^x(\alpha) &= \begin{bmatrix}
	1 & 0 & 0 \\
    0 & \cos(\alpha) & -\sin(\alpha) \\
    0 & \sin(\alpha) & \cos(\alpha)
	\end{bmatrix} \\
	R^y(\beta) &= \begin{bmatrix}
	\cos(\beta) & 0 & \sin(\beta) \\
    0 & 1 & 0 \\
    -\sin(\beta) & 0 & \cos(\beta)
    \end{bmatrix} \\
	R^z(\gamma) &= \begin{bmatrix}
	\cos(\gamma) & -\sin(\gamma) & 0 \\
    \sin(\gamma) & \cos(\gamma) & 0 \\
    0 & 0 & 1
    \end{bmatrix}.
\end{aligned}
\end{equation}
%\normalsize
%
The total rotation, $R_{ij} \in SO(3) \subset \mathbb{R}^{3 \times 3}$, is
\begin{equation} \label{eq:total_rot}
	R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})	 = R^z(\gamma_{ij})  R^y(\beta_{ij})  R^x(\alpha_{ij}).
%R(\alpha_{ij}, \beta_{ij}, \gamma_{ij}) =
%\begin{bmatrix}
%\cos(\beta) \cos(\gamma) & \cos(\gamma)\sin(\alpha)\sin(\beta)-\cos(\alpha) \sin(\gamma) & \cos(\alpha)\cos(\gamma)\sin(\beta)+ \sin(\alpha)\sin(\gamma) \\
%\cos(\beta) \sin(\gamma) & \cos(\alpha) \cos(\gamma)+\sin(\alpha) \sin(\beta) \sin(\gamma) & \cos(\alpha) \sin(\beta) \sin(\gamma)-\cos(\gamma) \sin(\alpha) \\
%-\sin(\beta) & \cos(\beta) \sin(\alpha) & \cos(\alpha) \cos(\beta))
%\end{bmatrix}
\end{equation}
%
%\eqref{eq:total_rot} corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).

These rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$ are used in formulating the matrix $H$ in \eqref{eq:H_to_R} for angular synchronization, or the matrix $S$ in \eqref{eq:vdm_S} for vector diffusion maps to compute the optimal rotations (and, in the case of vector diffusion maps, deduce optimal temporal orderings) for the images.
%
The three-dimensional optimal rotations $R_{1, est}, \dots, R_{m, est}$ then give us the corresponding translations and rotations of the images.
%
We multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
%
$R_{i,est}^T$ then gives the rotation matrix which registers image $i$.
%
From a rotation matrix $R$, we compute the corresponding Euler angles $\alpha$, $\beta$, and $\gamma$ using the following relationships
\begin{equation} \label{eq:angle_invert}
\begin{aligned}
R_{1,1} & = \cos(\beta)\cos(\gamma) &
R_{2,1} & = \cos(\beta)\sin(\gamma) \\
R_{3,1} & = -\sin(\beta) &
R_{3,2} & = \sin(\alpha)\cos(\beta) \\
R_{3,3} & = \cos(\alpha)\cos(\beta)  &
\end{aligned}
\end{equation}
%
Inverting \eqref{eq:angle_relations} yields the desired approximate translations and rotation, and we round optimal translations to the nearest integer to remove the need for interpolation when translating the images.

\subsection{The eigenvalue spectrum}
